\paperTitle

### 2014-05-18 doc/icml2014/directed.tex

  \caption{
    \label{fig:bottleneckExclusiveViews}
  (a) A bottleneck can be estimated consistently
  using tensor factorization \citep{anandkumar13tensor}.
  (b) An example of exclusive views for a bidependent subset $S$.
  }
\end{figure}

@@

We now show that if a subset of hidden variables $S$ has the exclusive views property,
then we can recover the marginal distribution $\Pr(\bh_S)$.
Consider any $S = \{i_1, \ldots, i_m\}$ with the exclusive views property. Let
  $x_{v_j}$ be an exclusive view for $h_{i_j}$ in $S$ and define $\sV
  = \{v_1, \ldots, v_m\}$. % be a set of exclusive views for the clique $S$.
By the exclusive views property,
the marginal over the observed variables $\Pr(\bx_\sV)$
factorizes according to the marginal over the hidden variables $\Pr(\bh_S)$
times the conditional moments:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  &= \sum_{\vh_S} \Pr(\Sh{S}) 
                    \Pr(x_{v_1} \given h_{i_1}) \cdots \Pr(x_{v_m} \given h_{i_m}) \\
  &= Z_{S}(\mOpp{v_1}{i_1},\dots,\mOpp{v_m}{i_m}) \\
  &= Z_{S}(\mOppAll),
\end{align*}
where $\mOppAll = \mOpp{v_1}{i_1} \otimes \cdots \otimes \mOpp{v_m}{i_m}$ is the tensor product of
all the conditional moments.
Vectorizing, we have that
$Z_S \in \Re^{k^m}$,
$M_\sV \in \Re^{d^m}$,
and $\mOppAll \in \Re^{d^m \times k^m}$.
Since each $\mOpp{v}{i}$ has full column rank $k$,
the tensor product $\mOppAll$ has full column rank $k^m$.
Succinctly, $\mO_\sV$ (which can be estimated directly from data)
is a linear function of $Z_S$ (what we seek to recover).
We can solve for the hidden marginals $Z_S$ simply by projecting $\mO_\sV$ onto the column
space of $\mOppAll$:
\begin{align*}
  Z_{S} &= \mO_\sV(\mOpp{v_1}{i_1}\pinv,\cdots,\mOpp{v_m}{i_m}\pinv).
\end{align*}

@@

\begin{algorithm}
  \caption{\LearnMarginals~(pseudoinverse)}
  \label{algo:learnMarginals}
  \begin{algorithmic}
    \REQUIRE Hidden subset $S = \{ h_{i_1}, \dots, h_{i_m} \}$ with exclusive views $\sV = \{ x_{v_1}, \dots, x_{v_m} \}$
    and conditional moments $\mOpp{v_j}{i_j} = \Pr(x_{v_i} \mid h_{i_j})$.
    \ENSURE Marginals $Z_S = \Pr(\bh_S)$.
      %\STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \dots, x_{v_m}\}$.
      \STATE Return $\mH_S \gets \mO_{\sV}( \mOpp{v_1}{i_i}\pinv, \dots, {\mOpp{v_m}{i_m}}\pinv )$.
  \end{algorithmic}
\end{algorithm}

@@

The bottleneck property allows recovery of conditional moments,
and the exclusive views property allows recovery of hidden marginals.
But we will now show that the latter property is in fact implied by the former
property for special sets of hidden variables, which we call bidependent
sets (in analogy with biconnected components),
in which conditioning on one variable does not break the set apart:
\begin{definition}[Bidependent set]
We say that a subset of nodes $S$ is \emph{bidependent} if
conditioned on any $a \in S$, there is an active trail between any other two nodes $b,c \in S$.
\end{definition}
Note that all cliques are bidependent, but bidependent sets can have more conditional independences
(e.g., $\{h_1,h_2,h_3\}$ in \figureref{bottleneckExclusiveViews}(b)).
This will be important for pseudolikelihood in \sectionref{pseudolikelihood}.

@@

% DONE: move here because not central to story
\paragraph{Remarks.}
Note that having only two independent views for each $h_i \in S$ is sufficient for condition (i)
  of the exclusive views property, while three is needed for condition (ii).
The bottleneck property (\definitionref{bottleneck} can also be relaxed if some cliques
  share parameters (see examples below).
%For example, in the directed grid model, we can recover the conditional moments $O$ from
  %$x^a_1, x^b_1$ and $x^a_2$ with $h_1$ as the bottleneck.
  %Therefore, $h_2, h_3$ and $h_4$
  %need not be bottlenecks, and we can omit the observations $x^b_2, x^b_3$ and $x^b_4$.

Our method extends naturally to the case in which the observed variables
  are real-valued $x_v \in \Re^d$, as long as the hidden variables remain discrete. 
In this setting, the conditional moments
  $\mOpp{v}{i} \eqdef \E(x_v \given h_i) \in \Re^{d \times k}$
  would no longer be distributions but general rank $k$ matrices.
  %for each observed variable $x_v$ and
  %hidden variable $h_i$.
%\algorithmref{learnMarginals} % and \algorithmref{directed} 
%still applies and allow us to recover clique marginals $Z_\sC$.
%  \todo{make this a footnote if running out of space}
% PL: don't understand this comment
%In this case, we only require that the hidden variables in distinct
  %cliques share parameters.

>>>>

  \caption{
    \label{fig:bottleneckExclusiveViews}
  (a) A bottleneck $h_1$ has three conditionally independent views $x_1,x_2,x_3$.
  %using tensor factorization \citep{anandkumar13tensor}.
  (b) A bidependent subset $S$ has exclusive views $\{x_1,x_2,x_3,x_4\}$.
  }
\end{figure}

@@

We now show that if a subset of hidden variables $S$ has the exclusive views property,
then we can recover the marginal distribution $\Pr(\bh_S)$.
Consider any $S = \{h_{i_1}, \ldots, h_{i_m}\}$ with the exclusive views property. Let
  $x_{v_j}$ be an exclusive view for $h_{i_j}$ in $S$ and define $\sV
  = \{x_{v_1}, \ldots, x_{v_m}\}$. % be a set of exclusive views for the clique $S$.
By the exclusive views property,
the marginal over the observed variables $\Pr(\bx_\sV)$
factorizes according to the marginal over the hidden variables $\Pr(\bh_S)$
times the conditional moments:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  &= \sum_{\vh_S} \Pr(\Sh{S}) 
                    \Pr(x_{v_1} \given h_{i_1}) \cdots \Pr(x_{v_m} \given h_{i_m}) \\
  &= Z_{S}(\mOpp{v_1}{i_1},\dots,\mOpp{v_m}{i_m}) \\
  &= Z_{S}(\mOppAll),
\end{align*}
where $\mOppAll = \mOpp{v_1}{i_1} \otimes \cdots \otimes \mOpp{v_m}{i_m}$ is the tensor product of
all the conditional moments.
Vectorizing, we have that
$Z_S \in \Re^{k^m}$,
$M_\sV \in \Re^{d^m}$,
and $\mOppAll \in \Re^{d^m \times k^m}$.
Since each $\mOpp{v}{i}$ has full column rank $k$,
the tensor product $\mOppAll$ has full column rank $k^m$.
Succinctly, $\mO_\sV$ (which can be estimated directly from data)
is a linear function of $Z_S$ (what we seek to recover).
We can solve for the hidden marginals $Z_S$ simply by multiplying $\mO_\sV$ by the pseudoinverse of
$\mOppAll$:
\begin{align*}
  Z_{S} &= \mO_\sV(\mOpp{v_1}{i_1}\pinv,\cdots,\mOpp{v_m}{i_m}\pinv).
\end{align*}

@@

\begin{algorithm}
  \caption{\LearnMarginals~(pseudoinverse)}
  \label{algo:learnMarginals}
  \begin{algorithmic}
    \REQUIRE Hidden subset $S = \{ h_{i_1}, \dots, h_{i_m} \}$ with exclusive views $\sV = \{ x_{v_1}, \dots, x_{v_m} \}$
    and conditional moments $\mOpp{v_j}{i_j} = \Pr(x_{v_j} \mid h_{i_j})$.
    \ENSURE Marginals $Z_S = \Pr(\bh_S)$.
      %\STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \dots, x_{v_m}\}$.
      \STATE Return $\mH_S \gets \mO_{\sV}( \mOpp{v_1}{i_1}\pinv, \dots, {\mOpp{v_m}{i_m}}\pinv )$.
  \end{algorithmic}
\end{algorithm}

@@

The bottleneck property allows recovery of conditional moments,
and the exclusive views property allows recovery of hidden marginals.
But we will now show that the latter property is in fact implied by the former
property for special sets of hidden variables, which we call bidependent
sets (in analogy with biconnected components),
in which conditioning on one variable does not break the set apart:
\begin{definition}[Bidependent set]
We say that a subset of nodes $S$ is \emph{bidependent} if
conditioned on any $a \in S$, there is an active trail between any other two nodes $b,c \in S$.
\end{definition}
Note that all cliques are bidependent, but bidependent sets can have more conditional independences
(e.g., $\{h_1,h_2,h_3\}$ in \figureref{bottleneckExclusiveViews}(b)).
This will be important in \sectionref{pseudolikelihood}.

@@

% DONE: move here because not central to story
\paragraph{Remarks.}
Note that having only two independent views for each $h_i \in S$ is sufficient for condition (i)
  of the exclusive views property, while three is needed for condition (ii).
The bottleneck property (\definitionref{bottleneck}) can also be relaxed if some cliques
  share parameters (see examples below).
%For example, in the directed grid model, we can recover the conditional moments $O$ from
  %$x^a_1, x^b_1$ and $x^a_2$ with $h_1$ as the bottleneck.
  %Therefore, $h_2, h_3$ and $h_4$
  %need not be bottlenecks, and we can omit the observations $x^b_2, x^b_3$ and $x^b_4$.

Our method extends naturally to the case in which the observed variables
  are real-valued ($x_v \in \Re^d$), as long as the hidden variables remain discrete. 
In this setting, the conditional moments
  $\mOpp{v}{i} \eqdef \E(x_v \given h_i) \in \Re^{d \times k}$
  would no longer be distributions but general rank $k$ matrices.
  %for each observed variable $x_v$ and
  %hidden variable $h_i$.
%\algorithmref{learnMarginals} % and \algorithmref{directed} 
%still applies and allow us to recover clique marginals $Z_\sC$.
%  \todo{make this a footnote if running out of space}
% PL: don't understand this comment
%In this case, we only require that the hidden variables in distinct
  %cliques share parameters.
